Visualizing demographic evolution using geographically inconsistent census data

### 2019-06-13 paper/discussion.tex
new paper version

\section{Discussion and limitations}

Our objective was to leverage a graph based data representation and
visualization methods for the exploration of geographically inconsistent
region-based data. While we successfully replicated and corroborated results
from the literature, this method still has significant limitations.

Removing the need for geographical harmonization greatly reduces the amount of
work necessary to explore demographic data, but the method still requires
consistent variables across the years. Matching the variables can be trivial for
some aspects (Age), but challenging for others (Income). The divulged income
ranges vary over time and the actual values change due to inflation. Some
variables were not considered in earlier censuses, such as Race in Canada, or
Hispanic population in the USA, hampering its use when they are available. Since
this is only a prototype, we matched few aspects, but a proper demographic
analysis would benefit from all available information.

While using one small map for each year leads to an easier visualization that
does not require interaction, it does not scale if more than five or six years
are considered. In this case, it might be interesting to replace the larger map
considering each year individually, along with a temporal control for
navigation. Indeed, including more years would likely lead  to a stronger
mixture of colors in the trajectory map, leading to a predominantly grey hue.

The limitation on the number of displayed clusters because of the limited number
of distinguishable colors was significant. While increasing the number of
clusters would further complicate an already complex analysis, it might be
warranted for some regions. Color is a fundamental and intuitive tool for
information representation that can be coherently used across different plots,
so we opted to use it, even if in a limited way. With eight colors, there was
overlap between some clusters, the relevance gradient, and the color
combination.

Another limitation is the lack of control on how much the geographical
information will impact the clustering result. While the adopted method met our
needs for this work, a configurable control would add another dimension to the
exploration, allowing for more intra-cluster variance to obtain more 'compact'
clusters. We explored changing the number of content based augmented edges, but
this proved to be unreliable and hard to interpret. The~\emph{ClustGeo}
method~\cite{Chavent2017} can be a viable option for this, allowing a graph
based input and a hierarchical output, combined using a single mixing
parameter. Alternatively, one could cluster the changes~\cite{bian2018survey}
instead of the stable states.

There are also technological limitations, such as memory use on the
visualization client. To allow for changes on the CTs over the years, we use a
geographic file that contains all possible intersections, which can grow rather
large if the original city was expansive and contained several CTs, like NYC or
LA. However, the most significant technological limitation relates to parameters
that are not immediately interactive, such as the clustering configuration.
Since the clustering is computationally expensive and performed on the server,
which allows for cached results, some changes can take a few minutes to be
considered, removing any possibility of a continuous exploration.

Indeed, the cognitive load on the user is already significant, as we compromised
simplicity for accuracy. While other works labelled the clusters, as 'young
urban', 'struggling', and so on~\cite{Delmelle2016,Delmelle2017}, we show the
statistical characteristics of the clusters, which are harder to interpret, as
the data may have subtle nuances that labels would otherwise hide. This also led
to a crowded interface, mitigated somewhat the use of pop-up panels and collapsible
sections. For some cities, especially if they are small and stable, the panels
can appear redundant, but each provide a different way to interact with the
information that can ease the exploration process for larger and dynamic cities.

>>>>

\section{Discussion and limitations}
\label{sec:discussion}

Our objective was to leverage a network based data representation and
visualisation methods for the exploration of geographically inconsistent
region-based data. While we successfully replicated and corroborated results
from the literature, this method still has significant limitations.

Removing the need for geographical harmonisation greatly reduces the amount of
work necessary to explore demographic data, but the method still requires
consistent variables across the years. Matching the variables can be trivial for
some aspects (Age), but challenging for others (Income). The divulged income
ranges vary over time and the actual values change due to inflation. Some
variables were not considered in earlier censuses, such as Race in Canada, or
Hispanic population in the USA, hampering its use when they are available. Since
this is only a prototype, we matched few aspects, but a proper demographic
analysis would benefit from all available information.

The limitation on the number of displayed clusters because of the limited number
of distinguishable colours was significant. While increasing the number of
clusters would further complicate an already complex analysis, it might be
warranted for some regions. Colour is a fundamental and intuitive tool for
information representation that can be coherently used across different plots,
so we opted to use it, even if in a limited way. With eight colours, there was
overlap between some clusters, the relevance gradient, and the colour
combination.

The cognitive load on the user is significant, as we compromised simplicity for
accuracy. While other works labelled the clusters, as 'young urban',
'struggling', and so on~\citep{Delmelle2016,Delmelle2017}, we show the
statistical characteristics of the clusters, which are harder to interpret, as
the data may have subtle nuances that labels would otherwise hide. This also led
to a crowded interface, mitigated somewhat the use of pop-up panels and
collapsible sections. For some cities, especially if they are small and stable,
the panels can appear redundant, but each provide a different way to interact
with the information that can ease the exploration process for larger and
dynamic cities.

### 2019-06-13 paper/intro.tex
new paper version

% General context + motivation
\IEEEPARstart{U}{rban} sciences are blooming thanks to a renewed interest in
understanding and improving the urban environment. Visual analytics is following
this trend, fueled by new public datasets that encompass progressively more of
our daily lives~\cite{Chen2017}. There is no shortage of methods to explore
mobility patterns~\cite{VonLandesberger2016}, social media~\cite{Chen2017},
traffic~\cite{chen2015survey}, and so on, providing experts, planners, policy
makers, and the general population with deeper insights about their cities.

These new datasets usually contain GPS coordinates for the records, leading to
\emph{point-based} data. Combined with the corresponding timestamps, this data
is easily suitable for longitudinal analysis. But most demographic datasets are
\emph{region-based}, where the measurements are associated with pre-defined
regions, not only for an additional level of privacy protection, but because
some measurements only make sense over a defined area. Census data is a classic
example of this format, with datasets available from 1790 onward for the
US~\cite{nhgis}. Despite this unmatched temporal availability, longitudinal
analyses of census data are often restricted in time, especially when smaller
tabulation areas are considered, such as census tracts (CT) or dissemination
areas, which evolve to reflect changes in population density, leading to
geographic inconsistencies across time, and the traditional time-series based
approach is no longer viable.

However, these analyses are necessary to understand the urban environment.
Indeed, two different regions can have similar average income for a given year,
while one is experiencing a process of economic improvement and the other one
impoverishment. A single snapshot cannot be used to identify gentrification,
migration, education changes, or any of the relevant processes that happen over
time.

To overcome these inconsistencies, the traditional approach is the
\emph{geographical harmonization} of the data, the interpolation of the
measurements into a common set of
regions\cite{Logan2014,Hallisey2017,Allen2018}, so that each variable can be
represented using time-series. This is laborious work that inevitably introduces
some amount of error~\cite{Logan2016}, even when additional data is
provided~\cite{eicher2001dasymetric}. Nevertheless, this step is considered
mandatory in the current literature: \emph{"(...) tract-by-tract comparison is
not possible unless data from 2000 is interpolated to 2010 boundaries
(...)"}~\cite{Dmowska2017}.

The main contribution of this application paper is an visualization-based
alternative to the geographical harmonization, a combination of established
graph based processing and information visualization techniques allowing
tract-by-tract comparison, the identification and visualization of patterns of
demographic evolution without geographic harmonization, effectively removing one
of the most challenging problems in longitudinal demographic analysis. We also
include illustrative scenarios and our prototype is available at
\url{http://uoft.me/piccard}, including more than forty regions in the US and
Canada. The source code is publicly available at
\url{https://github.com/fabioasdias/piccard}.

>>>>

% General context + motivation

Neighbourhoods have increasingly become a central concept in social research and
targets for social
policy~\citep{sampson2012great,galster2019making,stone2015urban,looker2015nation}.
To be sure, a focus on neighbourhoods extends to the formative period of the
modern social sciences~\citep{Abbott1997}. Recent interest has at least partly
been rekindled through newly available longitudinal demographic
datasets~\citep{Logan2014,nhgis}, convenient computational
tools~\citep{rey2018spatio}, and new sources of data~\citep{Poorthuis2018}.

These advances have made neighbourhood constructs more tractable to quantitative
analysis. Yet new challenges have also emerged, especially at the convergence of
research on neighbourhood effects and neighbourhood dynamics. Neighbourhood
effects research assumes knowledge about the nature and scope of "the
neighbourhood" that presumably shapes individual
outcomes~\citep{Kwan2018,Shelton2019}. Concurrently, researchers note that
neighbourhoods are not necessarily fixed containers in which other processes
occur, but themselves dynamically
evolve~\citep{Delmelle2017,Reades2019,li2018new}. The result is to open up key
assumptions about neighbourhoods for theoretical and empirical examination: how
do we appropriately define and compare neighbourhoods at a given time?; how do
we appropriately define and compare the temporal trajectories of
neighbourhoods?; and can we do both at once, "fully
interactionally"~\citep{Abbott1997}: classify neighbourhoods now based on where
they came from and where they are going?

In principle, much of the recent research is committed to the proposition that
neighbourhoods are open and evolving entities. Ironically, its empirical
practice tends to rely on methods that require fixed geographical regions. This
requirement is difficult to satisfy, as most longitudinal datasets are based on
pre-defined tabulation areas that are routinely modified by data collection
agencies, usually to follow population changes.

The standard approach then is to \emph{geographically harmonise} data. This
involves interpolating existing measurements into a common set of
regions~\citep{Logan2014,Hallisey2017,Allen2018}. Recent computational tools
have somewhat simplified this process~\citep{rey2018spatio}, but it still
involves non-trivial questions: which geometry to use as target, how to
apportion the variables, or how to combine data from different sources. Further,
these question do not necessarily have optimal answers. Indeed, regardless of
how well this process is performed, it still introduces
errors~\citep{Logan2016}, even when additional data is
provided~\citep{eicher2001dasymetric}. Essentially, harmonisation generates
artificial data points that can potentially lead to inaccurate results, even
though they are seldom interpreted as such. Nevertheless, because there has been
no viable alternative, and the results often appear plausible, these concerns
are generally overlooked. The result is that the harmonisation approach is
virtually mandatory in the current literature: \emph{"(...) tract-by-tract
comparison is not possible unless data from 2000 is interpolated to 2010
boundaries (...)"}~\citep{Dmowska2017}, \emph{"(...) This limits cross-year
comparison since data are not representative of the same spatial units. (...)"}
~\citep{Allen2018}.

The main contribution of this paper is a method for longitudinal data processing
that works with the original data by leveraging a network based representation.
It enables tract-by-tract comparison and the identification of patterns of
demographic evolution \emph{without geographic harmonisation}.

To allow a proper examination of our method and its results, we built an online
interactive system using this representation. It enables users to visualise,
interpret, and explore trajectories of neighbourhood change. This interface
helps validate our method, by allowing it to be compared to existing and future
methods. Further, it is a significant contribution to the research community: it
provides a vehicle for quickly and easily grasping complex long-term changes,
experimenting with different parameters to interactively learn from data, and
making neighbourhood change research publicly transparent. The interface thus
responds to increasing concerns about reproducibility and transparency, as well
as ongoing attention to the value of visualisation in scientific research and
communication.

We start by presenting an intuitive example of our representation in
Section~\ref{sec:introduction}, then we review the relevant literature on
longitudinal studies, data representation, clustering, and spatio-temporal
visualisation in Section~\ref{sec:related}. Our methodology is introduced in
detail in Section~\ref{sec:method}, along with the included interface.
Illustrative scenarios for Chicago, Toronto, and Los Angeles are presented in
Section~\ref{sec:study} and the feedback of five field experts are summarised in
Section~\ref{sec:expert}. Our prototype system is available at
\censure{\url{http://uoft.me/piccard}}, including more than forty regions in the
US and Canada. The source code is publicly available at
\censure{\url{https://github.com/fabioasdias/piccard}}.\footnote{The editors are
considering, at our request, an exception to the double-blind requirement to
allow access to the system. We provided them with the URLs of the system, code,
and documentation separately.}

\section{Intuition}
\label{sec:intuition}
While utterly simple, the network model breaks from the deeply rooted
traditional tabular paradigm in a significant way. Instead of requiring the data
as a collection of fixed entities which properties evolve over time, literally
rows in a table with temporal values as columns, it represents each measurement
as a separate entity and encodes the evolution of these entities over time.

To ease this cognitive transition, we start with an intuitive example of how it
works, using a small portion of a fictitious urban region illustrated on the
left part of Figure~\ref{fig:intuition}. This example includes three different
times ($t_0,t_1,t_2$), with different aggregation areas identified as letters
from A to H. For $t_0$, the initial time, we have areas A and B, with small
houses and a park, respectively. The park remains stable (B, E, and H), but the
houses are partially replaced by larger buildings (C and F).

This example also illustrates some of the challenges of harmonisation. None of
these aggregation areas is clearly suitable as an interpolation target. In fact,
adopting any of these areas as a target would require merging heterogeneous
regions and/or dividing homogeneous regions. For instance, by choosing the
regions of $t_1$, A would be split to match regions C and D, which appears to be
a rather reasonable approximation in this homogeneous artificial example, albeit
the fallacy of division. Region F would be similarly split to match C, but F
would be split and merged with G to match D, potentially leading to statistical
measurements that do not properly represent either region.

Real world measurements are seldom as homogeneous and noise-free as this
artificial example. By splitting and merging the data to fit arbitrary borders,
that were not necessarily coherent at the time of the measurement, harmonisation
increases the distance between measurement and reality.

\begin{figure}
    \centering 
    \subfloat{
        \includegraphics[width=0.475\linewidth]{intuition.png}
    }
    \subfloat{
        \begin{tikzpicture}
            \node (push) at (-2,0) {};
            \node [circle, draw=red, thick] (a) at (0,4) {A};
            \node [circle, draw=green, thick] (b) at (4,4) {B};
            \node [circle, draw=blue,  thick] (c) at (0,2) {C};
            \node [circle, draw=red, thick] (d) at (2,2) {D};
            \node [circle, draw=green, thick] (e) at (4,2) {E};
            \node [circle, draw=blue, thick] (f) at (0,0) {F};
            \node [circle, draw=red, thick] (g) at (2,0) {G};
            \node [circle, draw=green, thick] (h) at (4,0) {H};
            \draw (a) -- (c);
            \draw (a) -- (d);
            \draw (b) -- (e);
            \draw (c) -- (f);
            \draw (d) -- (f);
            \draw (d) -- (g);
            \draw (e) -- (h);
        \end{tikzpicture}
    } \caption{Network based spatio-temporal data representation. \textbf{Left}:
    Three temporal stages of the evolution of a fictitious urban area, with
    aggregation areas A to H. \textbf{Right}: Network representation of the
    aggregation areas where the colours identify similar regions.
        \label{fig:intuition}}
\end{figure}

Instead, we propose a network representation. Intuitively, a \emph{network}
(also called a \emph{graph}) is a collection of entities (nodes) that are
related to each other (edges). In this case, each different aggregation area is
represented as a node and we connect nodes that have overlapping geographical
areas, leading to the network illustrated on the right of
Figure~\ref{fig:intuition}. By partitioning the network into connected nodes
that are similar, we are effectively identifying clusters in the spatio-temporal
data, as illustrated by the colours of the nodes on the right side of
Figure~\ref{fig:intuition}. Further, the possible evolution paths can also be
obtained by computing temporal sequences of nodes, in this case: (A, C, F), (A,
D, F), (A, D, G), and (B, E, H). This representation is also suited for
geographically consistent regions, as illustrated by the stable park in this
example, and is therefore a generalisation of the traditional paradigm.

%I'm adding this because Jeff Allen (Farber's guy) struggled with this.
Note that the edges of this network merely encode that two regions are related.
This is a binary information, there is no apportionment, no areal measurements,
no population percentages associated with the edge. Indeed, our method also
connects regions of the same time that share borders, representing exactly that
they are neighbouring areas.
